---
title: "Data Visualizations"
format:
  html:
    code-fold: true
execute:
  warning: false
  message: false
---

[← Back to home](index.html){.back-home}


```{r, include=FALSE}
library(tidyverse)
library(patchwork)
library(praise)
```


# Eclipses

This data has information about the paths of the solar eclipses on October 14, 2023 and April 8, 2024 in the United States. The eclipse datasets were accessed via the TidyTuesday project (R for Data Science community), Week of April 9, 2024. The underlying data originate from NASA’s Scientific Visualization Studio.

```{r}
eclipse_annular_2023 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-09/eclipse_annular_2023.csv')
eclipse_total_2024 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-09/eclipse_total_2024.csv')
eclipse_partial_2023 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-09/eclipse_partial_2023.csv')
eclipse_partial_2024 <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-04-09/eclipse_partial_2024.csv')

```

```{r, fig.width=15, fig.height=5}
# tidy 2024 data
tot_24 <- eclipse_total_2024 %>% 
  mutate(time = eclipse_6 - eclipse_1, eclipse = "total") %>% 
  select(state, name, lat, lon, time, eclipse)

part_24 <- eclipse_partial_2024 %>% 
  mutate(time = eclipse_5 - eclipse_1, eclipse = "partial") %>% 
  select(state, name, lat, lon, time, eclipse)

eclipse_24 <- rbind(tot_24, part_24) %>% 
  mutate(time = as.numeric(time))

# plot 2024 data
library(ggnewscale)
states <- map_data("state")

twenty_four <- ggplot(states) +
  geom_polygon(fill = "white", color = "black",
               aes(long,lat, group = group)) +
  geom_point(data = filter(eclipse_24, lat < 51 & lat > 24 & lon < 0 & eclipse == "total"), 
             aes(x = lon, y= lat, color = time), size = 0.001) +
  scale_colour_gradientn(colors = c("blue", "turquoise")) +
  labs(color = "time of overlap (sec)\n total eclipse") +
  new_scale_color() +
  geom_point(data = filter(eclipse_24, lat < 51 & lat > 24 & lon < 0 & eclipse == "partial"), 
             aes(x = lon, y= lat, color = time), size = 0.001) +
  scale_colour_gradientn(colors = c("red", "pink")) +
  ggtitle("2024 Solar Eclipse") +
  labs(x = "", y = "", color = "time of overlap (sec)\n partial eclipse")
  


# tidy 2023 data
tot_23 <- eclipse_annular_2023 %>% 
  mutate(time = eclipse_6 - eclipse_1, eclipse = "total") %>% 
  select(state, name, lat, lon, time, eclipse)

part_23 <- eclipse_partial_2023 %>% 
  mutate(time = eclipse_5 - eclipse_1, eclipse = "partial") %>% 
  select(state, name, lat, lon, time, eclipse)

eclipse_23 <- rbind(tot_23, part_23) %>% 
  mutate(time = as.numeric(time))


# plot 2023 data
states <- map_data("state")

twenty_three <- ggplot(states) +
  geom_polygon(fill = "white", color = "black",
               aes(long,lat, group = group)) +
  geom_point(data = filter(eclipse_23, lat < 51 & lat > 24 & lon < 0 & eclipse == "total"), 
             aes(x = lon, y= lat, color = time), size = 0.25) +
  scale_colour_gradientn(colors = c("blue", "turquoise")) +
  labs(color = "time of overlap (sec)\n total eclipse") +
  new_scale_color() +
  geom_point(data = filter(eclipse_23, lat < 51 & lat > 24 & lon < 0 & eclipse == "partial"), 
             aes(x = lon, y= lat, color = time), size = 0.001) +
  scale_colour_gradientn(colors = c("red", "pink")) +
  ggtitle("2023 Solar Eclipse") +
  labs(x = "", y = "", color = "time of overlap (sec)\n partial eclipse")
  

# put plots together
twenty_three + twenty_four + plot_layout(ncol = 2)
```

# Mt. Vesuvius

This dataset contains information about seismic events detected at Mount Vesuvius in Italy. The data was accessed via TidyTuesday, Week of May 13, 2025, based on data from the Italian Istituto Nazionale di Geofisica e Vulcanologia (INGV). The raw data was saved as individual CSV files from the GOSSIP website and some values were translated from Italian to English.

```{r}
vesuvius <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-13/vesuvius.csv')

```

```{r, fig.width=12, fig.height=5}
library(sf)
library(ggspatial)

# drop NA lat/long since we're plotting on a map
vesuvius <- vesuvius |>
  drop_na(latitude, longitude)


# convert data to be compatible with sf
vesuvius_sf <- st_as_sf(vesuvius, coords = c("longitude", "latitude"), crs = 4326)


# map for energy release
magnitude <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = duration_magnitude_md), size = 1, alpha = 0.6) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +  
  labs(title = "Energy Release",
       x = "Longitude", y = "Latitude")


# map for year
year <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = year), size = 1, alpha = 0.6) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +  
  labs(title = "Year",
       x = "Longitude", y = "Latitude")


# map for depth of seismic event
depth <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = depth_km), size = 1, alpha = 0.6) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +
  labs(title = "Depth",
       x = "Longitude", y = "Latitude")

# put them all together
magnitude + year + depth + 
  plot_layout(ncol = 2) + plot_annotation("Earthquakes Around Mt. Vesuvius", caption = "Energy release measured by 'duration magnitude' (Md)")
```

# H.E.L.P. (Health Evaluation and Linkage to Primary Care)

The HELP dataset comes from a longitudinal study of adults experiencing housing instability and substance use disorders in Boston in the early 2000's, designed to evaluate factors influencing linkage to medical and social services after enrollment. It includes detailed measures of incarceration history, substance use, mental health (including CES-D depression scores), and time to successful linkage to care. The dataset can be accessed [here.](https://melindahiggins2000.github.io/N736/helpdata.html)

```{r}
library(wesanderson)
library(mosaic)
library(circlepackeR)

# data
HELPfull <- read.csv("HELPdata_2025.csv")

HELPfull <- HELPfull |>
  mutate(dayslink = as.numeric(dayslink),
         linkstatus = as.numeric(linkstatus),
         group = as.factor(group))


# bin the variables to "create" hierarchy

# each bin 19 months
HELPfull$jail_bin <- cut(HELPfull$jail_mos, breaks = 3, include.lowest = TRUE, labels = c("Low Jail Time", "Medium Jail Time", "Higher Jail Time"))

# each bin 14 years
HELPfull$mj_bin  <- cut(HELPfull$h10_lt, breaks = 3, include.lowest = TRUE, labels = c("Low Marijuna Use", "Medium Marijuana Use", "Higher Marijuana Use"))

# each bin about 20 points
HELPfull$dep_bin  <- cut(HELPfull$ces_d, breaks = 3, include.lowest = TRUE, labels = c("Low Depression Score", "Medium Depression Score", "Higher Depression Score"))

# create data with subgroups
circle_data <- HELPfull |>
  group_by(jail_bin, mj_bin, dep_bin) %>%
  summarise(value = mean(dayslink, na.rm = TRUE), .groups = "drop")

library(data.tree)

# circlepackeR requires hierarchy
circle_data$pathString <- paste("world", circle_data$jail_bin, circle_data$dep_bin, circle_data$mj_bin, sep = "/")

tree_data <- as.Node(circle_data)


library(htmltools)

# plot
p <- circlepackeR(tree_data, size = "value", 
             color_min = "#91BAB6", 
             color_max = "#F03801")

tagList(tags$h3(style = "color:#000000; font-family:Ipsum Larem; font-weight:400;",
          "Circle Pack of Jail Time, Depression Score, and Marijuana Use"),
  p)
```

When we stratify the cohort by incarceration exposure, depressive symptom burden, and marijuana-use history, the average time to linkage differs across subgroups—suggesting these risk/complexity dimensions may be associated with slower (or faster) linkage.

