---
title: "Mt. Vesuvius"
format: html
editor: source
execute: 
  warning: false
  message: false

project:
  type: website
  output-dir: docs
---

## the data

This dataset contains information about seismic events detected at Mount Vesuvius in Italy. It comes from the Italian Istituto Nazionale di Geofisica e Vulcanologia (INGV)'s Data Portal and can be explored along with other seismic areas on the GOSSIP website. The raw data was saved as individual CSV files from the GOSSIP website and some values were translated from Italian to English

```{r}
vesuvius <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-05-13/vesuvius.csv')

```
## earthquake maps

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(patchwork)
library(praise)
```

```{r}
# drop NA lat/long since we're plotting on a map
vesuvius <- vesuvius |>
  drop_na(latitude, longitude)
```

```{r}
# convert data to be compatible with sf
vesuvius_sf <- st_as_sf(vesuvius, coords = c("longitude", "latitude"), crs = 4326)

```

```{r}
# map for energy release
magnitude <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = duration_magnitude_md), size = 1) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +  
  labs(title = "Energy Release",
       x = "Longitude", y = "Latitude")
```

```{r}
# map for year
year <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = year), size = 1) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +  
  labs(title = "Year",
       x = "Longitude", y = "Latitude")
```


```{r}
# map for depth of seismic event
depth <- vesuvius_sf |>
  ggplot() +
  geom_sf(aes(color = depth_km), size = 1) +
  scale_color_gradient(low = "#3B4CC0", high = "#FADA48") +
  theme_minimal() +
  theme(axis.text.x = element_text(vjust = 0.5, hjust=1, size = 5),
        axis.title.x = element_text(vjust = -3),
        axis.text.y = element_text(size = 5)) +
  labs(title = "Depth",
       x = "Longitude", y = "Latitude")
```


```{r, fig.width=12, fig.height=5}
# put them all together
magnitude + year + depth + 
  plot_layout(ncol = 2) + plot_annotation("Earthquakes Around Mt. Vesuvius", caption = "Energy release measured by 'duration magnitude' (Md)")
```

## Depth vs. Duration Magnitude (Md) by Year

```{r}
vesuvius |>
  ggplot(aes(x = duration_magnitude_md, y = depth_km)) +
  geom_point(aes(color = year)) +
  facet_wrap(~year)
```

```{r}
praise()
```



